---
title: "An introduction to the SpatialFeatures Package"
author:
- name: Guan Gui, Shila Ghazanfar
  affiliation:
  - School of Mathematics and Statistics, The University of Sydney, Sydney, NSW, 2006, Australia
  - Charles Perkins Centre, The University of Sydney, Sydney, NSW, 2006, Australia
  - Sydney Precision Data Science Centre, The University of Sydney, Sydney, NSW, 2006, Australia
output:
  BiocStyle::html_document:
    toc_float: true
  BiocStyle::pdf_document: default
package: SpatialFeatures
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Getting started: SpatialFeatures}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

# SpatialFeatures
The R package `SpatialFeatures` contains functions to extract features from molecule-resolved spatial transcriptomics data using objects from the `MoleculeExperiment` class.

# Demonstrating Major Functions and Visualizing SpatialFeatures

## Load Required Libraries
```{r}
library(SpatialFeatures)
library(MoleculeExperiment)
library(ggplot2)
library(SummarizedExperiment)
library(dplyr)
```

## Create Example Dataset
Load an example `MoleculeExperiment` object with Xenium data:
```{r}
repoDir <- system.file("extdata", package = "MoleculeExperiment")
repoDir <- paste0(repoDir, "/xenium_V1_FF_Mouse_Brain")

me <- readXenium(repoDir, keepCols = "essential")
me
```

## Generate `SpatialFeature` Data Based on Different Segmentation Methods

This function `GenerateFeatureData(me)` processes the given data (`me`) and returns feature data based on multiple assay types, including:
- Sub-Sector
- Sub-Concentric
- Super-Sector
- Super-Concentric
```{r}
feature_data_list <- SpatialFeatures::GenerateFeatureData(me)
lapply(feature_data_list, head, n = 4)
```


## Load boundaries to `MoleculeExperiment` object  

This function `loadBoundariesToMEforAll(me)` divides the image data into equal areas and processes each area internally to extract the respective boundary features, updating the `MoleculeExperiment` object with these new features.
```{r warning=FALSE}
boundaries_me_list <- SpatialFeatures::loadBoundariesToMEforAll(me)
me_4 = boundaries_me_list[[4]]
me_4
```

## Draw the Feature Boundary Plots

Visualize the spatial molecule and cell boundary map with SpatialFeature boundaries.
```{r}
ggplot_me() +
  # add molecule points and colour by feature name
  geom_point_me(me_4, byColour = "feature_id", size = 0.1) +
  # add nuclei segments and colour the border with red
  geom_polygon_me(me_4, assayName = "sub_sector", fill = NA, colour = "red") +
  # add cell segments and colour by cell id
  geom_polygon_me(me_4, byFill = "segment_id", colour = "black", alpha = 0.1) +
  # zoom in to selected patch area
  coord_cartesian(xlim = c(4875, 4950), ylim = c(6385, 6455))

ggplot_me() +
  # add molecule points and colour by feature name
  geom_point_me(me_4, byColour = "feature_id", size = 0.1) +
  # add nuclei segments and colour the border with red
  geom_polygon_me(me_4, assayName = "sub_concentric", fill = NA, colour = "red") +
  # add cell segments and colour by cell id
  geom_polygon_me(me_4, byFill = "segment_id", colour = "black", alpha = 0.1) +
  # zoom in to selected patch area
  coord_cartesian(xlim = c(4875, 4950), ylim = c(6385, 6455))

ggplot_me() +
  # add molecule points and colour by feature name
  geom_point_me(me_4, byColour = "feature_id", size = 0.1) +
  # add nuclei segments and colour the border with red
  geom_polygon_me(me_4, assayName = "super_sector", fill = NA, colour = "blue") +
  # add cell segments and colour by cell id
  geom_polygon_me(me_4, byFill = "segment_id", colour = "black", alpha = 0.1) +
  # zoom in to selected patch area
  coord_cartesian(xlim = c(4875, 4950), ylim = c(6385, 6455))

ggplot_me() +
  # add molecule points and colour by feature name
  geom_point_me(me_4, byColour = "feature_id", size = 0.1) +
  # add nuclei segments and colour the border with red
  geom_polygon_me(me_4, assayName = "super_concentric", fill = NA, colour = "blue") +
  # add cell segments and colour by cell id
  geom_polygon_me(me_4, byFill = "segment_id", colour = "black", alpha = 0.1) +
  # zoom in to selected patch area
  coord_cartesian(xlim = c(4875, 4950), ylim = c(6385, 6455))
```

## Create Entropy Matrix

This function `EntropyMatrix()` computes the entropy of a counts matrix from a `MoleculeExperiment` object based on the given assay type.
```{r}
em_list_4 = SpatialFeatures::EntropyMatrix(me_4, c("sub_sector", "sub_concentric", "super_sector", "super_concentric"), nCores = 1)
lapply(em_list_4, head, n = 4)
```

## Create `SummarizedExperiment` Object

This function `EntropySummarizedExperiment()` Convert the entropy matrix into a `SummarizedExperiment` object.
```{r}
se_4 = SpatialFeatures::EntropySummarizedExperiment(em_list_4, me_4)
se_4
```

# Full Process Tutorial: From `MoleculeExperiment` to `SummarizedExperiment`
This section provides a complete tutorial on transforming an input `MoleculeExperiment` (me) object into a final output `SummarizedExperiment` (se) object.

## Read the Original MoleculeExperiment Object
Load the example data from the `MoleculeExperiment` package:
```{r}
repoDir <- system.file("extdata", package = "MoleculeExperiment")
repoDir <- paste0(repoDir, "/xenium_V1_FF_Mouse_Brain")

me <- readXenium(repoDir, keepCols = "essential")
```

## Load `SpatialFeatures` Boundaries
Load the `SpatialFeatures` boundaries into the `MoleculeExperiment` object:
```{r}
melist <- loadBoundariesToMEforAll(me, k = 5)
```

## Compute and Combine the Entropy Matrices
Initialize an Empty List to Store Split `SummarizedExperiment` Objects. Loop through the desired range of r values, compute the entropy matrices, and combine them into a single `SummarizedExperiment` object:
```{r}
se_list <- list()

for (r in 1:12) {
  if (melist[r] == "null") next
  
  splitted_me <- melist[[r]]
  em_list <- SpatialFeatures::EntropyMatrix(splitted_me, c("sub_sector", "sub_concentric", "super_sector", "super_concentric"), nCores = 1)
  se_object <- SpatialFeatures::EntropySummarizedExperiment(em_list, splitted_me)
  se_list[[r]] <- se_object
}

se_list <- Filter(Negate(is.null), se_list)
combined_se <- SpatialFeatures::combineSE(se_list)
```

## Inspect the Combined SummarizedExperiment Object
Check the structure and content of the combined SummarizedExperiment object:
```{r}
combined_se
colData(combined_se)
rowData(combined_se)
```

